InputFinder: Reverse Engineering Closed Binaries via Side Channels

### 2015-05-17 paper/main.tex
final draft of the paper

\documentclass{acm_proc_article-sp}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage{algorithm}

@@

\title{InputFinder: Reverse Engineering Closed Binaries via Side Channels}
\numberofauthors{2}
\author{
\alignauthor
Bogdan Copos\\
	\affaddr{University of California, Davis}\\
	\email{bcopos@ucdavis.edu}
\alignauthor
Praveen Murthy\\
	\affaddr{Fujitsu Laboratories of America}\\
	\email{praveen.murthy@us.fujitsu.com}
}
\maketitle

\begin{abstract}
The effectiveness of many dynamic program analysis techniques depends heavily on the completeness of the test suite applied during the analysis process.
Test suites are often composed by developers and aim at testing all of the functionality of a software system.
However, test suites may not be complete, if they exist at all.
To date, only two methods exist for automatically generating test input for closed binaries: fuzzing and symbolic execution.
Despite previous successes of these methods in identifying bugs, both techniques have limitations.
In this paper, we propose a new method for autonomously generating valid input and identifying protocols for closed x86 binaries.
To assess its effectiveness, we have developed \tool{} and tested it against binaries from the DARPA Cyber Grand Challenge example set.
Our evaluations show that ...
\end{abstract}

@@

%motivation
Without complete test suites, the program cannot be properly tested and bugs may be omitted during testing.
While valid input can be learned from software documentation or source code, when such information is not available or complete, other methods must be used.
Automated dynamic analysis techniques such as fuzzing and symbolic execution attempt to generate input for a given program.
Currently these are the only two options available and both approaches present limitations.
Fuzzing done randomly with no inference attempted of the program's behavior cannot be effective as the input space is usually too large.
Symbolic execution has traditionally been used with manual intervention to identify potential symbolic variables to begin exploration with, a luxury we do not have when analyzing closed binaries autonomously, in addition to scalability issues and limitations of decision procedures that can be used to solve the constraints.

In this paper, we introduce a new method, called InputFinder, for generating valid user-input for closed binaries. 
This new method could be useful in autonomous vulnerability scanning systems, or for reverse engineering unknown binaries.
Our method is composed of two main components.
The first component exploits side channels to build valid input for closed binaries.
Specifically, as the program is given various input, \tool{} records the number of instructions retired during the program's execution for each input and uses a statistical measure to infer the program's expected input.
The number of instructions retired differs from the number of instructions executed by a processing unit.
Because of the out-of-order CPU pipeline, a CPU may execute more instructions than necessary (e.g. as a result of wrong branch prediction).
The instructions retired represent the subset of instructions executed which leave the \textit{Retirement Unit} once execution has been deemed correct (i.e. the instructions which actually impact the program).
The same side channel is used by \tool{} to also find the expected input size and to categorize the expected input based on type.
The second component uses the discovered input to build a protocol state machine associated with the tested program.
This enables the generation of a more thorough test suite.

The main contributions of this paper are as follows:
\begin{itemize}
	\item We introduce a new approach to automatically generate input valid strings for unknown, closed-source binaries by leveraging hardware instruction counters. Our method relies on observing changes in the number of instructions retired by programs during their input validation process to learn valid input strings.
	\item We describe how our method can be used to not only find valid input strings, but also to determine the expect input size and classify input based on type (e.g. alphabetic, alphanumeric, numeric, etc.).
	\item We develop a method for generating protocol state machines using the valid input previously found.
	\item We implement the techniques presented as a proof of concept tool and evaluate it using binaries published by DARPA as part of the 2014 Cyber Grand Challenge competition \cite{darpacgc}. Additionally, we compare our approach to the manual process used by a security specialist to crack a binary as part of a job interview \cite{interviewbinary}.
\end{itemize}

The paper is organized as follows.
Section \ref{background} covers background information relating to concepts utilized in this work.
Section \ref{prevwork} discusses related works and highlights differences between the related works and the work presented here.
The details of our approach are explained in Section \ref{methodology}. First, the method for finding valid input is described. Then we introduce our method for building protocol state machines using the discovered input.
Section \ref{results} describes evaluation of the work presented and the results.
In Section \ref{futurework}, we share future plans for our work and explain cases which the proposed method does not handle.

\section{Background} \label{background}
\subsection{Side Channels} \label{bg_sidechannels}
Side channels are streams of information that can be retrieved from the hardware of a device running a given program.
This information can be used to gather knowledge regarding the program's internal components.
Side channels have been widely explored by security researchers in order to discover weaknesses in programs \cite{weinbergside, schindler2002combined, genkin2014rsa, genkinstealing}.
In cryptography, side channel attacks exploit such information channels to defeat crypto systems \cite{zhou2005side, black2002black, okeya2006side}.
Recently, side channel attacks, such as OpenSSL's CREAM cache timing attack have made headlines \cite{creamssl}.
Similar to CREAM and other side channel attacks, the technique discussed in this paper exploits such information channels to generate valid input for binaries.
\subsection{Hardware counter registers} \label{bg_hardwarereg}
Most modern microprocessors are equipped with a set of special-purpose registers used to count hardware events.
Each register, or counter, can be programmed to measure specific events such as cache misses, floating point operations, and even instructions retired.
Such hardware performance counters are often used by software developers to analyze their programs and improve the running time and efficiency.
However, in this paper we show how such performance counters can also be exploited as side channels.
Utilities, such as Linux/Unix \textit{perf}, take advantage of such hardware counters in order to provide developers with useful information about a program's performance.
The \textit{perf} utility allows developers to specify the desired hardware event they wish to measure.
Once events have been selected, \textit{perf} interacts with the appropriate kernel modules to program the hardware counters accordingly.
In this work, we use \textit{perf} to count the number of \textbf{user-land} instructions retired as a program is tested with various input.
\textbf{User-land} instructions are instructions of a program which run outside the operating system's kernel.
While other options are available for counting instructions retired, we choose \textit{perf} for its efficiency and minimal overhead.
The number of \textit{total} instructions retired varies significantly for a binary from execution to execution \cite{weaver2008can}.
This phenomenon reflects fluctuations in instructions retired by the kernel depending on the state of the machine due to cache-misses and other factors.
Since we rely on this count, such variations are problematic.
However, experimentally, the number of \textit{user-land} instructions only varies, on average, by one instruction, giving us a reliable counter.
\subsection{Execution/Dynamic Matching} \label{bg_dynamicsim}
In many application areas, including clone detection and software debugging, there is a need for comparing executions of a program (or multiple versions of a program).
One of the proposed approaches aimed at tackling this issue is dynamic matching.
Dynamic matching works by first collecting execution traces of different executions.
Such traces can be collected through a number of ways.
One option includes the use of a tool such as \textit{Pin} \cite{pintool}.
\textit{Pin} is a framework for building customized binary analysis tools and can be used to design a tool which creates a transcript of the instructions retired by a program.
Once the traces are collected, they are analyze and mappings between instructions of different executions are produced.
These mappings can be used to understand the relationship between two executions (or two versions of a program).
In this work, we rely on the concept of dynamic matching and the \textit{Pin} framework to develop a protocol state machine generator for closed binaries.

@@

Symbolic execution is another method used for generating valid input.
When a program analyzer symbolically executes a program, it does not use concrete values.
Instead, the input is represented as a mathematical formula composed of symbols and equations resulting from the constraints (related to the input) gathered during the program's execution.
Specifically, the tester first marks some memory locations as symbolic.
This can also be done automatically by, for example, marking the buffer used in \textit{read()} system call as symbolic.
However, this does not always produce good results, especially if the program reads one character at a time and then copies the read character into a second buffer.
The program is then executed and the tester enters some input.
Regardless of the value entered, the program replaces the input with a symbol.
As the input flows through the program and encounters conditional statements, the symbolic value of the input changes to reflect the faced constraints.
At the end of the execution, a constraint solver takes the formulat representing the symbolic value of the input and attempts to solve it in order to obtain valid program input.
Symbolic execution has been proven effective by previous works \cite{symbolicexecution, exesymex}.
Similar to white box fuzzing, symbolic execution has a fundamental limitation with respect to constraint solving.
While symbolic execution engines perform well with linear constraints, non-linear constraints greatly impact efficiency.
One of the most important works in this area is described in \cite{exesymex} which introduces EXE, a dynamic analysis tool capable of generating inputs that crash programs by running the programs with symbolic input.
Unlike the EXE symbolic execution engine, the SAGE tool and SmartFuzz, the approach presented in our work is not hindered by constraint solvers.
Additionally, compared to EXE, our method does not require any instrumentation or source code.
\tool{} does not execute the program with symbolic input but rather it bootstraps itself by building up a set of valid inputs (from an initially empty set) by gathering information about the execution of a program.

White box fuzzing leverages the symbolic execution traces of a given binary to produce more inputs.
One of the many previous works is \cite{godefroid2008automated} which introduces SAGE, an automated white box fuzzer.
White box fuzzing differs from black box fuzzing in that it uses the program's feedback to make inferences about valid input and uses constraint solvers to generate more input.
White box fuzzing can attain better code coverage but can be slow and is crippled by innate limitations of constraint solvers.
The only similarity between the approach introduced in this paper and white box fuzzing is that both approaches use information from the program's execution to craft valid input.
However, our method does not use symbolic execution.

One of the works utilizing white box fuzzing is \cite{molnar2009dynamic} which introduces SmartFuzz, a dynamic test generation tool.
SmartFuzz uses symbolic execution to learn about the program's input and passes that information to a fuzzer for test case generation.
The main difference between our work and SmartFuzz is that our work does not use symbolic execution to gain insight into the format of the input.
Similar to SmartFuzz, the output of our work can be used by a fuzzer to generate test cases.

Another related work is presented in \cite{smartfuzzer}, where the authors use a combination of static and dynamic analysis to build a smart fuzzer.
Their work uses static analysis to collect preliminary information such as an approximation of the inter procedural control flow graph.
Having gathered static information, their dynamic analysis engine executes the program while monitoring dependencies between the input and the dynamic control flow path experienced by the program.
Their work also takes advantage of a constraint solver that, given a path of the program, generates input which forces the program to follow the chosen path.
Our work \tool{} differs from \cite{smartfuzzer} in several ways.
One difference is that our work uses neither static analysis nor constraint solvers.
Instead we rely only on information extracted from the execution of the uninstrumented program.

There have also been numerous works studying and applying dynamic similarity.
In \cite{blanketexec}, the authors propose blanket execution, a novel dynamic equivalence testing technique.
Blanket execution is designed to collect information about the execution of functions, including side effects, which is then used for similarity testing.
Another important work is \cite{differential} which introduces \textit{differential slicing} for the purpose of identifying flow and value differences across multiple executions of the same program for security applications.
Both our approach and the previous approaches mentioned above use a dynamic approach to gather information about the program.
However, our approach aims at finding differences in executions of a given program in order to make inferences about the underlying protocol expected by the program.
This allows us to discover differences in inputs which experience different control flow paths.

Another area of research relevant to this paper is protocol reverse engineering.
An important result in this area is \cite{prospex} which introduces a technique for extracting protocol specifications from packet captures and automatically generating input which can be used by a stateful fuzzer.
There are two notable difference between Prospex and our work.
While the end result is similar, \tool{} assumes no prior knowledge about the tested program and hence it does not utilize packet captures for protocol specification extraction.
Instead, \tool{} infers the protocol through dynamic analysis by automatically finding input and using that to generate a protocol state machine.
To date, we do not know of any protocol reverse engineering techniques which work without any prior knowledge about the program's input.

@@

\tool{} is composed of two main modules.
The first module is responsible for discovering input and properties of the expected input, including input size and the type of the input.
The input is then be used by the second component to generate a protocol state machine.
The protocol state machine can be used by third party tools such as stateful fuzzers for even better code coverage.
\subsection{Finding Valid Input}
The majority of software programs accept input, perform transformations on the input and output results.
However, most programs do not accept completely random input.
Input usually passes through an input validation filter.
Such filters are snippets of code responsible for distinguishing between valid or good input (i.e. input the program was constructed to understand and accept) and bad input (i.e. input which is not useful or does not follow the desired format).
Although validation mechanisms can be very sophisticated, they are often a combination of string comparisons and conditional statements.
Our method exploits changes in the number of user-land instructions retired during multiple executions with varying inputs to make inferences about the program's validation mechanism.
These changes reflect different execution paths of the program as a response to both valid and invalid input.
Assuming most input is invalid, our method observes differences in the number of instructions retired to detect input which passes the validation mechanism.

\begin{figure*}[t]
\centering
\includegraphics[height=2.5in,width=6.5in]{architecture.png}
\caption{This figure depicts the architecture of the finding input process. There are three main parts: the execution engine, the valid character filter and the input string builder.}
\label{fig:architecture}
\end{figure*}

@@

% \algref{findinputcode}{line}
The steps of the process for finding valid input are described in algorithm \ref{findinputcode} and outlined below. Starting with an empty string, \textit{s}:
\begin{enumerate}
\item \tool{} executes the program once for every printable character and records the number of user-land instructions retired by the program with the given input. The input is a string concatenation of the string \textit{s} and the current printable character. The Unix \textit{perf} utility is used for monitoring the number of instructions retired. Note that \tool{} does not record the total number of instructions but rather user-land instructions. The reason for this, as explained in Section \ref{bg_hardwarereg}, is that the total number of instructions retired can vary greatly from execution to execution due to operating system factors (e.g. branch prediction, cache). The Expect tool is used for automating the program interaction. This step is shown in line \ref{execute} of algorithm \ref{findinputcode}
\item Once all executions have been completed (one for each character), the mode number of instructions retired is computed across all executions. The mode is defined as the value that appears most often in the set of recordings. This is done in line \ref{mode} of algorithm \ref{findinputcode}.
\item Next is the filtering stage which is responsible for identifying characters likely to be part of a valid input string at the current index. Under the assumption that most characters are not valid for the current index, \tool{} gathers all characters with the number of instructions retired outside the range of the mode +/- an epsilon value, as shown in lines \ref{filterbegin}-\ref{filterend} of algorithm \ref{findinputcode}. These characters represent valid characters for the current index of the input string. The epsilon value accounts for small variations in the hardware counters. While the initial reaction is to expect a valid character to result in more instructions retired than a non-valid character, this is not always the case. As discussed earlier, the input validation mechanism can be complex and can vary in behavior. The validation mechanism may verify input against all accepted input strings before denying the provided input. Such a scenario results in more instructions retired for invalid input than valid input.
\item For each valid character \textit{c}, \tool{} forks, with every child process repeating steps 1-4, starting with $ c + s $ as the initial input string, where $+$ signifies concatenation (starting at line \ref{fork} of algorithm \ref{findinputcode}). If no valid characters are found, the end of a valid input string has been reached and the string is recorded. If the current index is 0 and no valid characters have been found, the given binary has no predefined input commands.
\end{enumerate}

\begin{figure*}[t]
\centering
\includegraphics[height=2in,width=6.5in]{string_builder.png}
\caption{This diagram visualizes the input string being constructed throughout five consecutive time stages of the execution of 
\tool{}. Multiple outgoing arrows represent a process forking. This occurs when multiple valid characters are detected for a given index of the input string and, as a result, the finding input process forks to concurrently handle all of the discovered characters}
\label{fig:string_builder}
\end{figure*}

Figure~\ref{fig:string_builder} depicts the results of the above process over the span of five consecutive time stages.
As seen in the figure, at time \textit{T0}, \tool{} begins with an empty string.
At time \textit{T1}, \tool{} finds valid characters for index 0 of the input string and is forked into \textit{n} child processes where \textit{n} is the number of valid characters found.
This continues until all child processes are done.
% TODO: snippets of code of input validation code from our experiments (also for finding input size)

\subsubsection{Finding Desired Input Size}
The same side channels can be used to detect the expected size of the program's input or even the size of the program's input buffer.
Specifically, there is a change in the behavior of the program with respect to the number of instructions retired when the size of the input is larger than expected.
Most programs are designed to read a predefined number of bytes from the source of input and store the read bytes in a buffer.
A popular way of doing that is by using the \textit{fgets()} or \textit{scanf()} functions of the GNU C stdio library.
As the input size grows, the number of instructions retired also increases linearly.
This happens because additional instructions are retired to read more characters and store them into the buffer.
However, when the input string is larger than the expected size the number of instructions retired no longer increases as a function of the input size.
This behavior can be observed and exploited to find the expected input size, which may often also be the size of the input buffer.
Our method observes such differences in the number of user-land instructions retired as the program is given increasingly larger input.

To start, \tool{} finds the set of all invalid characters for index 0 of all input strings and picks one at random.
The character selected is used to build invalid input strings of various sizes.
It is crucial to use invalid input to avoid noisy fluctuations in the number of instructions retired due to valid input processing.
Initially, the program is retired with short length input strings (i.e. two and three characters).
The number of user-land instructions retired during these two executions are recorded.
These two values may be equal or differ, depending on the program's logic.
The two values may be equal if, for example, the program uses the system call \textit{read()} directly to read data from the input file descriptor.
Since we are measuring the number of user-land instructions retired (not also instructions retired by the kernel), the instructions retired by the \textit{read} functions are not included in our counter so there will be no linear increase in the number of instructions retired as the input size is increased.
If the two values are equal, the program is executed with increasingly large input until a change in the number of executions is recorded due to the program logic.
There may not always be a change (depending on the program logic when the input size exceeds the expected size), however, we note that this scenario is not very interesting since the expected size of the input may also be obtained by executing the program with \textit{strace}.
If the values differ (as discussed earlier), the difference in the number of instructions retired is recorded and the program is executed, increasing the size of the input, until a change is detected.
Instead of increasing the input size linearly, the input size is increased exponentially until a change is detected.
Once the change is detected, the input size is decreased to find the exact size at which the change happens.
It is also possible that the number of instructions retired oscillates between two (or more values).
One example is a palindrome finder.
The same underlying concept can be applied to such cases.

>>>>

\documentclass{acm_proc_article-sp}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage{url}
\usepackage{listings}
\usepackage{color}

@@

\title{InputFinder: Reverse Engineering Closed Binaries via Side Channels}
%\numberofauthors{2}
%\author{
%\alignauthor
%Bogdan Copos\\
%	\affaddr{University of California, Davis}\\
%	\email{bcopos@ucdavis.edu}
%\alignauthor
%Praveen Murthy\\
%	\affaddr{Fujitsu Laboratories of America}\\
%	\email{praveen.murthy@us.fujitsu.com}
%}
\maketitle

\begin{abstract}
The effectiveness of many dynamic program analysis techniques depends heavily on the completeness of the test suite applied during the analysis process.
Test suites are often composed by developers and aim at testing all of the functionality of a software system.
However, test suites may not be complete, if they exist at all.
To date, only two methods exist for automatically generating test input for closed binaries: fuzzing and symbolic execution.
Despite previous successes of these methods in identifying bugs, both techniques have limitations.
In this paper, we propose a new method for autonomously generating valid input and identifying protocols for closed x86 binaries.
To assess its effectiveness, we have developed \tool{} and tested it against binaries from the DARPA Cyber Grand Challenge example set.
Our evaluations show that our method is not only effective in finding input and determining whether a protocol is expected but can also find unexpected control flow paths.
\end{abstract}

@@

%motivation
Without complete test suites, the program cannot be properly tested and some bugs may be omitted during testing.
While valid input can be learned from software documentation or source code, when such information is not available or complete, other methods must be used.
Automated dynamic analysis techniques such as fuzzing and symbolic execution attempt to generate input for a given program.
Currently these are the only two options available and both approaches present limitations.
Fuzzing done randomly with no inference attempted of the program's behavior cannot be effective as the input space is usually too large.
Symbolic execution has traditionally been used with manual intervention to identify potential symbolic variables to begin exploration with, a luxury we do not have when analyzing closed binaries autonomously, in addition to scalability issues and limitations of decision procedures that are used to solve the constraints.

In this paper, we introduce a new method, called InputFinder, for generating valid user-input for closed binaries. 
This new method could be useful in autonomous vulnerability scanning systems, or for reverse engineering unknown binaries.
Our method is composed of two main components.
The first component exploits side channels to build valid input for closed binaries.
Specifically, as the program is given various inputs, \tool{} records the number of instructions \textbf{retired} during the program's execution for each input and uses a statistical measure to infer the program's expected input.
The number of instructions retired differs from the number of instructions executed by a processing unit.
Because of the out-of-order CPU pipeline, a CPU may execute more instructions than necessary (e.g. as a result of wrong branch prediction).
The instructions retired represent the subset of instructions executed which leave the \textit{Retirement Unit} once execution has been deemed correct (i.e. the instructions which actually impact the program).
The same side channel is used by \tool{} to also find the expected input size and to categorize the expected input based on type.
The second component uses the discovered input to build a protocol state machine associated with the tested program.
This enables the generation of a more thorough test suite.

The main contributions of this paper are as follows:
\begin{itemize}
	\item We introduce a new approach to automatically generate input valid strings for unknown, closed-source binaries by leveraging hardware instruction counters. Our method relies on observing changes in the number of instructions retired by programs during their input validation process to learn valid input strings.
	\item We describe how our method can be used not only to find valid input strings, but also to determine the expect input size and classify input based on type (e.g. alphabetic, alphanumeric, numeric, etc.).
	\item We develop a method for generating protocol state machines using the valid input previously found.
	\item We implement the techniques presented as a proof of concept tool and evaluate it using binaries published by DARPA as part of the 2014 Cyber Grand Challenge competition \cite{darpacgc}. Additionally, we compare our approach to the manual process used by a security specialist to crack a binary as part of a job interview \cite{interviewbinary}.
\end{itemize}

The paper is organized as follows.
Section \ref{prevwork} discusses related works and highlights differences between the related works and the work presented here.
Section \ref{background} covers background information relating to concepts utilized in this work.
The details of our approach are explained in Section \ref{methodology}. First, the method for finding valid input is described. Then we introduce our method for building protocol state machines using the discovered input.
Section \ref{results} describes the evaluation of the work presented and the results.
In Section \ref{futurework}, we share future plans for our work and explain cases which the proposed method does not handle.

@@

Symbolic execution is another method used for generating valid input.
When a program analyzer symbolically executes a program, it does not use concrete values.
Instead, the input is represented as a mathematical formula composed of symbols and equations resulting from the constraints (related to the input) gathered during the program's execution.
Specifically, the tester first marks the input as symbolic.
This can also be done automatically via taint analysis.
The program is then executed by an interpreter that assumes symbolic values for those symbolic input locations.
As the program executes, whenever operations are performed on those symbolic variables, a formula in terms of those symbols keeps track of the mathematical operations performed on the symbolic variables.
The formula can also have constraints on those symbols based on outcomes of conditional branches encountered during the executions.
At the end of execution, or during certain portions of the execution, the formula is examined by a constraint solver to determine whether it is satisfiable (meaning that a solution exists to make the formula true), and if it does, concrete values for all symbolic variables in the formula are computed.
Those concrete values will then constitute a set of test cases that can execute the program along the set of paths (because of branches) encountered in the formula.
Symbolic execution has been proven effective by previous works \cite{symbolicexecution, exesymex}.
One of the most important works in this area is described in \cite{exesymex} which introduces EXE, a dynamic analysis tool capable of generating inputs that crash programs by running the programs with symbolic input.

White box fuzzing leverages the symbolic execution traces of a given binary to produce more inputs.
One of the many previous works is \cite{godefroid2008automated} which introduces SAGE, an automated white box fuzzer.
White box fuzzing differs from black box fuzzing in that it uses the concrete inputs to gather constraints and then systematically negates constraint clauses to generate additional inputs that cover more branches in the program.
White box fuzzing can attain better code coverage but can be slow.
Symbolic execution (and consequently white box fuzzing) is crippled by innate limitations of constraint solvers.
While symbolic execution engines perform well with linear constraints, non-linear constraints greatly impact efficiency.
One of the works utilizing white box fuzzing is \cite{molnar2009dynamic} which introduces SmartFuzz, a dynamic test generation tool.
SmartFuzz uses symbolic execution to learn about the program's input and passes that information to a fuzzer for test case generation.

The main difference between our work and SmartFuzz is that our work does not use symbolic execution to gain insight into the format of the input.
Similar to SmartFuzz, the output of our work can be used by a fuzzer to generate test cases.
Unlike the EXE symbolic execution engine, the SAGE tool, and SmartFuzz, the approach presented in our work is not hindered by constraint solvers.
Additionally, compared to EXE, our method does not require any instrumentation or source code.
\tool{} does not execute the program with symbolic input but rather it bootstraps itself by building up a set of valid inputs (from an initially empty set) by gathering information about the execution of a program.

Another related work is presented in \cite{smartfuzzer}, where the authors use a combination of static and dynamic analysis to build a smart fuzzer.
Their work uses static analysis to collect preliminary information such as an approximation of the inter procedural control flow graph.
Having gathered static information, their dynamic analysis engine executes the program while monitoring dependencies between the input and the dynamic control flow path experienced by the program.
Their work also takes advantage of a constraint solver that, given a path of the program, generates input which forces the program to follow the chosen path.
Our work \tool{} differs from their approach in several ways.
One difference is that our work uses neither static analysis nor constraint solvers.
Instead our input finding method relies only on information extracted from the execution of the un-instrumented program.

There have also been numerous works studying and applying dynamic similarity.
In \cite{blanketexec}, the authors propose blanket execution, a novel dynamic equivalence testing technique.
Blanket execution is designed to collect information about the execution of functions, including side effects, which is then used for similarity testing.
Another important work is \cite{differential} which introduces \textit{differential slicing} for the purpose of identifying flow and value differences across multiple executions of the same program for security applications.
Both our approach and the previous approaches mentioned above use a dynamic approach to gather information about the program.
However, our approach aims at finding differences in executions of a given program in order to make inferences about the underlying protocol expected by the program.
This allows us to discover the orders of inputs which experience unique control flow paths.

Another area of research relevant to this paper is protocol reverse engineering.
An important result in this area is \cite{prospex} which introduces a technique for extracting protocol specifications from packet captures and automatically generating input which can be used by a stateful fuzzer.
There are two notable differences between Prospex and our work.
While the end result is similar, \tool{} assumes no prior knowledge about the tested program and hence it does not utilize packet captures for protocol specification extraction.
Instead, \tool{} infers the protocol through dynamic analysis by automatically finding input and using that to generate a protocol state machine.
Caballero et. al introduce Polyglot \cite{polyglot}, a tool that uses dynamic binary analysis to observe how programs process the received application data and uses that information to infer the protocol's message format.
Their work is similar because of the use of dynamic binary analysis and the use of information from program traces to make inferences about the protocol format.
However, our work does not try to determine the protocol message format but rather the high level protocol.
Additionally, unlike Polyglot, our work does not use test cases or previously known input.
To date, we do not know of any protocol reverse engineering techniques which work without any prior knowledge about the program's input.

\section{Background} \label{background}

\subsection{Side Channels} \label{bg_sidechannels}
Side channels are streams of information that can be retrieved from the hardware of a device running a given program.
This information can be used to gather knowledge regarding the program's internal components.
Side channels have been widely explored by security researchers in order to discover weaknesses in programs \cite{weinbergside, schindler2002combined, genkin2014rsa, genkinstealing}.
In cryptography, side channel attacks exploit such information channels to defeat crypto systems \cite{zhou2005side, black2002black, okeya2006side}.
Recently, side channel attacks, such as OpenSSL's CREAM cache timing attack have made headlines \cite{creamssl}.
Similar to CREAM and other side channel attacks, the technique discussed in this paper exploits such information channels to generate valid input for binaries.
\subsection{Hardware counter registers} \label{bg_hardwarereg}
Most modern microprocessors are equipped with a set of special-purpose registers used to count hardware events.
Each register, or counter, can be programmed to measure specific events such as cache misses, floating point operations, and even instructions retired.
Such hardware performance counters are often used by software developers to analyze their programs and improve the running time and efficiency.
However, in this paper we show how such performance counters can also be exploited as side channels.
Utilities, such as Linux/Unix \textit{perf}, take advantage of such hardware counters in order to provide developers with useful information about a program's performance.
The \textit{perf} utility allows developers to specify the desired hardware event they wish to measure.
Once events have been selected, \textit{perf} interacts with the appropriate kernel modules to program the hardware counters accordingly.
In this work, we use \textit{perf} to count the number of \textbf{user-land} instructions retired as a program is tested with various input.
\textbf{User-land} instructions are instructions of a program which run outside the operating system's kernel.
While other options are available for counting instructions retired, we choose \textit{perf} for its efficiency and minimal overhead.
The number of \textit{total} instructions retired varies significantly for a binary from execution to execution \cite{weaver2008can}.
This phenomenon reflects fluctuations in instructions retired by the kernel depending on the state of the machine due to cache-misses and other factors.
Since we rely on this count, such variations are problematic.
However, experimentally, the number of \textit{user-land} instructions only varies, on average, by one instruction, giving us a reliable counter.
\subsection{Execution/Dynamic Matching} \label{bg_dynamicsim}
In many application areas, including clone detection and software debugging, there is a need for comparing executions of a program (or multiple versions of a program).
One of the proposed approaches aimed at tackling this issue is dynamic matching.
Dynamic matching works by first collecting execution traces of different executions.
Such traces can be collected through a number of ways.
One option includes the use of a tool such as \textit{Pin} \cite{pintool}.
\textit{Pin} is a framework for building customized binary analysis tools and can be used to design a tool which creates a transcript of the instructions retired by a program.
Once the traces are collected, they are analyze and mappings between instructions of different executions are produced.
These mappings can be used to understand the relationship between two executions (or two versions of a program).
In this work, we rely on the concept of dynamic matching and the \textit{Pin} framework to develop a protocol state machine generator for closed binaries.

@@

\tool{} is composed of two main modules.
The first module is responsible for discovering input and properties of the expected input, including input size and the type of the input.
The input generated by the first module is then used by the second module to generate a protocol state machine.
The protocol state machine can be used by third party tools such as stateful fuzzers for even better code coverage.
\subsection{Finding Valid Input}
The majority of software programs accept input, perform transformations on the input, and output results.
However, most programs do not accept completely random input.
Input usually passes through an input validation filter.
Such filters are snippets of code responsible for distinguishing between valid or good input (i.e. input the program was constructed to understand and accept) and bad input (i.e. input which is not useful or does not follow the desired format).
Although validation mechanisms can be very sophisticated, they are often a combination of string comparisons and conditional statements.
Code snippets of input validation from our tested binaries can be found in Appendix A.
Our method exploits changes in the number of user-land instructions retired during multiple executions with varying inputs to make inferences about the program's validation mechanism.
These changes reflect different execution paths of the program as a response to both valid and invalid input.
Assuming most input is invalid, our method observes differences in the number of instructions retired to detect input which passes the validation mechanism.

\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{architecture.pdf}
\caption{This figure depicts the architecture of the finding input process. There are three main parts: the input string builder, the performance counter (i.e. \textit{perf} utility), and the valid character filter}
\label{fig:architecture}
\end{figure*}

@@

% \algref{findinputcode}{line}
The steps of the process for finding valid input are described in algorithm \ref{findinputcode} and outlined below. Starting with an empty string, \textit{s}:
\begin{enumerate}
\item \tool{} executes the program once for every printable character and records the number of user-land instructions retired by the program with the given input. The input is a string concatenation of the string \textit{s} and the current printable character. The Unix \textit{perf} utility is used for monitoring the number of instructions retired. Note that \tool{} does not record the total number of instructions but rather user-land instructions. The reason for this, as explained in Section \ref{bg_hardwarereg}, is in contrast to user-land instructions, that the total number of instructions retired can vary greatly from execution to execution for various reasons (e.g. branch prediction, cache). The Expect tool is used for automating the program interaction. This step is shown in line \ref{execute} of algorithm \ref{findinputcode}.
\item Once all executions have been completed (one for each character), the mode number of instructions retired is computed across all executions. The mode is defined as the value that appears most often in the set of recordings. This is done in line \ref{mode} of algorithm \ref{findinputcode}.
\item Next is the filtering stage which is responsible for identifying characters likely to be part of a valid input string at the current index. Under the assumption that most characters are not valid for the current index, \tool{} gathers all characters with the number of instructions retired outside the range of the mode +/- an epsilon value, as shown in lines \ref{filterbegin}-\ref{filterend} of algorithm \ref{findinputcode}. These characters represent valid characters for the current index of the input string. The epsilon value accounts for small variations in the hardware counters. While the initial reaction is to expect a valid character to result in more instructions retired than a non-valid character, this is not always the case. As discussed earlier, the input validation mechanism can be complex and can vary in behavior. The validation mechanism may verify input against all accepted input strings before denying the provided input. Such a scenario results in more instructions retired for invalid input than valid input.
\item For each valid character \textit{c}, \tool{} forks, with every child process repeating steps 1-4, starting with $ c + s $ as the initial input string, where $+$ signifies concatenation (starting at line \ref{fork} of algorithm \ref{findinputcode}). If no valid characters are found, the end of a valid input string has been reached and the string is recorded. If the current index is 0 and no valid characters have been found, the given binary has no predefined input commands.
\end{enumerate}

\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{string_builder.pdf}
\caption{This diagram visualizes the input strings being constructed throughout five consecutive time stages of the execution of 
\tool{}. Multiple outgoing arrows represent a process forking. This occurs when multiple valid characters are detected for a given index of the input string and, as a result, the finding input process forks to concurrently handle all of the discovered characters.}
\label{fig:string_builder}
\end{figure*}

Figure~\ref{fig:string_builder} depicts the results of the above process over the span of five consecutive time stages.
As seen in the figure, at time \textit{T0}, \tool{} begins with an empty string.
At time \textit{T1}, \tool{} finds valid characters for index 0 of the input string and is forked into \textit{n} child processes where \textit{n} is the number of valid characters found.
This continues until all child processes are done.

\subsubsection{Finding Desired Input Size}
The same side channels can be used to detect the expected size of the program's input or even the size of the program's input buffer.
Specifically, there is a change in the behavior of the program with respect to the number of instructions retired when the size of the input is larger than expected.
Most programs are designed to read a predefined number of bytes from the source of input and store the read bytes in a buffer.
A popular way of doing that is by using the \textit{fgets()} or \textit{scanf()} functions of the GNU C stdio library.
As the input size grows, the number of instructions retired also increases linearly.
This happens because additional instructions are needed to read more characters from the input source.
However, when the input string is larger than the expected size, the number of instructions retired no longer increases as a function of the input size.
This behavior can be observed and exploited to find the expected input size, which may often also be the size of the input buffer.
Our method observes such differences in the number of user-land instructions retired as the program is given increasingly larger input.

To start, \tool{} finds the set of all invalid characters for index 0 of all input strings and picks one at random.
The character selected is used to build invalid input strings of various sizes.
It is crucial to use invalid input to avoid noisy fluctuations in the number of instructions retired due to valid input processing.
Initially, the program is executed twice: once with an input string of two invalid characters and once with an input string of three invalid characters.
The number of user-land instructions retired during these two executions are recorded.
These two values may be equal or different, depending on the program's logic.
The two values may be equal if, for example, the program uses the system call \textit{read()} directly to read data from the input file descriptor.
Since we are measuring the number of user-land instructions retired (not also instructions retired by the kernel), the instructions retired by the \textit{read()} function are not included in our counter so there will be no linear increase in the number of instructions retired as the input size is increased.
If the two values are equal, the program is executed with increasingly large input until a change in the number of executions is recorded due to the program logic.
If the values differ (as discussed earlier), the difference in the number of instructions retired is recorded and the program is executed, increasing the size of the input, until a change is detected.
Instead of increasing the input size linearly, the input size is increased exponentially until a change is detected.
Once the change is detected, the input size is decreased to find the exact size at which the change happens.
It is also possible that the number of instructions retired oscillates between two (or more values).
One example is a palindrome finder.
The same underlying concept can be applied to such cases.

Determining the exact size is useful when performing symbolic execution on a buffer where one cannot make the size of the buffer symbolic (since frameworks such as \textit{Fuzzball} do not allow for that).
In such cases, being able to determine the buffer size is necessary and useful to instruct the symbolic execution engine the exact size of the buffer that should be made symbolic.
We detected 5 crash inputs in the tested binaries because of our finding input size approach.
